---
title: "BIOL668_FINAL"
author: "Natalia Contreras"
date: "2025-05-12"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

From meeting notes:
* Use "all_no_O_P_S.csv" containing source samples from freshwater, gut and wastewater. 
* Use "TJ_metagenomemodified.csv" containing sink samples from the different sites of the Tijuana River
* merge "all_no_O_P_S.csv" + "TJ_metagenomemodified.csv"
* Use map_no_O_P_S.csv"as the mapping file 
* Perform cluster analysis (I tried PCA and perferred NMDS)
* Look at which groups are close/far away
* Which samples within a certain centroid of a group are close to sink samples?
* assess centroid by eye for now (still need to do)
* Which are the top 10 closest? (still need to do)
* make plots 


```{r libraries}
library(tidyverse)
library(vegan)
library(ggfortify) 
library(factoextra)
library(ggforce)
#install.packages("ggforce")
```

```{r load data }
#load in: source, sink, and metadata
#file containing source information
site_source <- read.csv("all_no_O_P_S.csv", row.names = 1, check.names = FALSE)

#actual Tijuana River sample data
tj_sink <- read.csv("TJ_metagenomemodified.csv", row.names = 1, check.names = FALSE)

#metaadata containing all sample information
metadat <- read.csv("map_no_O_P_S.csv")

```


```{r data pre-processing}
#select for common taxa
common_taxa <- intersect(rownames(site_source), rownames(tj_sink))
source_filter <- site_source[common_taxa, ]
tj_filter <- tj_sink[common_taxa, ]


#merge into one single dataframe with matching rows and columns
otu_combined <- cbind(source_filter, tj_filter)
otu_tranpose <- as.data.frame(t(otu_combined)) #to be properly used for analysis

#replace zeros with very small numbers
otu_tranpose[otu_tranpose == 0] <- 0.001

#normalize data by applying central log ratio
clr <- function(x) sweep(log(x), 1, rowMeans(log(x)), "-") #creates a clr function

#applied clr and creates dataframe 
otu_clr <- as.data.frame(clr(otu_tranpose))

```

# PCA 
Combined source and sink dataframe, CLR transformed. 

```{r PCA}
PCA <- prcomp(otu_clr, center = TRUE, scale. = TRUE)
summary(PCA)
```
```{r scree plot}
#visualize which pca loadings explain most variance
screeplot(PCA, type = "lines", main = "Scree Plot")
```
```{r autoplot}
#biplot
autoplot(PCA, data = otu_clr,
         loadings = FALSE, loadings.label = FALSE) +
  theme_minimal() +
  labs(title = "PCA Biplot of CLR-Transformed OTU Data")

#plot by color 
#extracts PCA scores and adds metadata
pca_df <- PCA$x %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(metadat, by = c("SampleID" = "Sample"))

#pca_df, shows tj samples as NA
pca_df$Site[is.na(pca_df$Site)] <- "TJ" #treats NAs as actual tijuana river samples

#plots by site
ggplot(pca_df, aes(PC1, PC2, color = Site))  + geom_point(size = 2) +
  theme_minimal() +
  labs(title = "PCA of CLR-Transformed OTU Data",
       x = "PC1", y = "PC2") #NAs are the Tijuana river samples 

```



# NMDS Analysis

```{r NMDS}
#distance matrix calculation using euclidean
otu_dist <- dist(otu_clr, method = "euclidean")

#run NMDS analysis
nmds_res <- metaMDS(otu_dist, k = 2, trymax = 100)

#extrats scores and creates a new dataframe
nmds_scores <- as.data.frame(scores(nmds_res))
nmds_scores$SampleID <- rownames(nmds_scores)

#adds metadata
nmds_scores <- left_join(nmds_scores, metadat, by = c("SampleID" = "Sample"))


```
```{r}
#NAs treated as the acutal "TJ" samples 
nmds_scores$Site[is.na(nmds_scores$Site)] <- "TJ"

#clean for plotting by selecting specific columns 
nmds_scores_clean <- nmds_scores %>% select(NMDS1, NMDS2, SampleID, Site, Number) #selects NMDS1, NMDS2, SampleID, Site, Number to plot (makes a cleaner df, reorders)

```
```{r NMDS ggplot}
#I was trying to replicate the NMDS plot from the original paper 
ggplot(nmds_scores_clean, aes(x = NMDS1, y = NMDS2, color = Site)) +
  geom_point(size = 3) +
  geom_mark_ellipse(aes(group = Site), alpha = 0.1, show.legend = FALSE) +  
  theme_minimal() +
  labs(title = "NMDS of Source Metagenomes",
       x = "NMDS1", y = "NMDS2") +
  scale_color_manual(values = c(
    "Freshwater" = "#19A52F",
    "Gut" = "#B31212",
    "Ocean" = "#10A9E3",
    "Pacific" = "#68E9E4",
    "Soil" = "#432E19",
    "Wastewater" = "#A9810F",
    "TJ" = "blue"  #sink sample
  )) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold"))
```
Note:
I attempted to perform the centroid distances and including the top 10 sources but I was unsuccessful. 
I followed these sources but didn't include my code since it led me nowhere. 

https://gis.stackexchange.com/questions/6025/finding-centroid-of-cluster-of-points-using-r 
https://search.r-project.org/CRAN/refmans/rearrr/html/centroid.html
https://stackoverflow.com/questions/64105486/how-to-find-an-x-amount-of-closest-elements-to-a-centroid
https://gis.stackexchange.com/questions/447437/finding-the-nearest-point-to-the-centroid-of-the-polygons
